Density-dependent selection and the limits of relative fitness

### 2018-04-26 manuscript.tex
l

[I'm going to revise this after your next round of comments]
Selection is commonly described by assigning relative fitness values to genotypes. Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses. Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’. Our model clearly distinguishes between density-dependent selection and changes in density driven by selection. These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium. Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly. We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density. Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

\noindent (191 words)

>>>>

Selection is commonly described by assigning relative fitness values to genotypes. Yet many populations are density-regulated, and the strength of selection can then depend on density. Additionally, selection can induce changes in density. When these two effects occur together, the resulting selection dynamics are not describable by density-independent relative fitness values, particularly if selection is strong. In most previous models of density-dependent selection, these effects are assumed to occur together (e.g. ``$K$-selection’’), and as a result relative fitness models seem to require weak selection. Here we argue that constant relative fitness models are valid under a substantially broader range of conditions. We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability. Our model contains a finite ``reproductive excess’’, which allows selection on competitive ability to occur without affecting density. Remarkably, we find that even strong selection is compatible with density-independent relative fitnesses provided that only one trait evolves at a time. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (187 words)
